---
title: 'Lecture 10: Statistics in hockey/soccer'
author: "Skidmore College"
output: beamer_presentation
fontsize: 10pt
---

```{r knitr_options , include=FALSE}
#Here's some preamble, which makes ensures your figures aren't too big
knitr::opts_chunk$set(fig.width=3.5, fig.height=2.3, warning=FALSE,
message=FALSE)
```

## Goals

- Multiple logistic regression

- Score effects

- Expected goals

- Advanced shot mapping

## Set-up: 

\scriptsize

NHL shot data

```{r}
library(RCurl); library(tidyverse)
githubURL <- "https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/pbp_data_hockey.rds"
pbp_data <- readRDS(gzcon(url(githubURL)))
names(pbp_data)
```

## Adjusted coordinates

\scriptsize

```{r}
pbp_data <- pbp_data %>% 
  mutate(coords_x_adj = ifelse(event_team == home_team, 
                               -1*abs(coords_x), abs(coords_x)), 
         coords_y_adj = ifelse(event_team == home_team & coords_x < 0, 
                               coords_y, -1*coords_y), 
         coords_y_adj = ifelse(event_team == away_team & coords_x > 0, 
                               coords_y, -1*coords_y))

```


## Player metrics

\footnotesize

```{r}
season_2018 <- pbp_data %>% 
  filter(season == 20172018) %>% 
  group_by(event_player_1, season) %>% 
  summarise(n_goals_18 = sum(event_type == "GOAL"), 
            n_xGs_18 = sum(shot_prob), 
            n_shots_18 = n()) %>% 
  filter(n_shots_18 >= 100) %>% 
  select(-season)

season_2019 <- pbp_data %>% 
  filter(season == 20182019) %>% 
  group_by(event_player_1, season) %>% 
  summarise(n_goals_19 = sum(event_type == "GOAL"), 
            n_xGs_19 = sum(shot_prob), 
            n_shots_19 = n()) %>% 
  filter(n_shots_19 >= 100) %>% 
  select(-season)
```

## Player metrics

\scriptsize

```{r}
season_combine <- season_2018 %>% inner_join(season_2019)
head(season_combine)
library(corrplot)
```

## Player metrics

```{r, fig.width = 3.5, fig.height = 3.2}
cor_players <- cor(season_combine[,2:7])
corrplot(cor_players, method = "number")
```

## Player metrics, conclusions



## Score effects

\footnotesize

```{r}
pbp_data <- pbp_data %>% 
  mutate(score_diff = ifelse(event_team == home_team, 
                             home_score - away_score, 
                             away_score - home_score), 
         score_diff_cat = case_when(score_diff <= -1 ~ "Down", 
                                    score_diff == 0 ~ "Tied", 
                                    score_diff >= 1 ~ "Up"), 
         is_goal = event_type == "GOAL")

pbp_data %>% 
  group_by(score_diff_cat) %>% 
  summarise(ave_goal = mean(is_goal), 
            ave_distance = mean(event_distance, na.rm = TRUE), 
            ave_Xg = mean(shot_prob))
```

## Score effects, conclusions

## Soccer data


\scriptsize

```{r}
library(RCurl)
library(tidyverse)
url <- getURL("https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/sb_shot_data.csv")
wwc_shot <- read.csv(text = url)
names(wwc_shot)
```

## Soccer data


\scriptsize

```{r}
wwc_shot_summary <- wwc_shot %>% 
  group_by(match_id, possession_team.name) %>% 
  summarise(n_goals = sum(shot.outcome.name == "Goal"))
wwc_shot_summary %>% 
  head()
```



## Expected goals/link to future

Summarize: *Expected goals 2.0* [http://www.11tegen11.com/2014/08/07/expected-goals-2-0-some-light-in-the-black-box/](http://www.11tegen11.com/2014/08/07/expected-goals-2-0-some-light-in-the-black-box/)

Summarize: *Best predictor of future performance is expected goals* [http://11tegen11.net/2015/01/05/the-best-predictor-for-future-performance-is-expected-goals/](http://11tegen11.net/2015/01/05/the-best-predictor-for-future-performance-is-expected-goals/)



## Expected goals, repeatability of finishing skill

Summarize: *Repeatability of finishing skill* [https://statsbomb.com/2017/07/quantifying-finishing-skill/](https://statsbomb.com/2017/07/quantifying-finishing-skill/)



## Randomness and expected goals

Summarize: *12 shots good, 2 shots better* [(http://thepowerofgoals.blogspot.com/2014/02/twelve-shots-good-two-shots-better.html)](http://thepowerofgoals.blogspot.com/2014/02/twelve-shots-good-two-shots-better.html)


## Expected goals and addition

Summarize: *Expected goals don't add*
[https://medium.com/@dannypage/expected-goals-just-don-t-add-up-they-also-multiply-1dfd9b52c7d0](https://medium.com/@dannypage/expected-goals-just-don-t-add-up-they-also-multiply-1dfd9b52c7d0)
