---
title: 'Lecture 5: Logistic regression & NFL kickers'
author: "Skidmore College"
output: beamer_presentation
fontsize: 10pt
---

## Preamble:


\footnotesize

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
nfl_kick <- read.csv("https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/nfl_fg.csv")
head(nfl_kick)
```

## Warm-Ups 1/2

- Identify the longest field goal kicked by each kicker

- Identify the rate of successful field goals in each season

## Warm ups 3/4

- Surfaces with Grass == `FALSE` occur on turf. What is the rate of field goals made on each surface?

- Identify the rate of successful field goals kicked between 48 and 52 yards


## Review: multivariate linear regression

Model: 

$y_i = \beta_0 + \beta_1*x_{i1} + \beta_2*x_{i2} + \ldots + \beta_{p-1}*x_{i,p-1} + \epsilon_i$


Assumptions: 

- $\epsilon_i \sim N(0, \sigma^2)$
- $\epsilon_i$,$\epsilon_{i'}$ independent for all $i, i'$
- Linear relationship between $y$ and $x$

## Example: NFL kickers

\footnotesize

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
nfl_kick <- read.csv("https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/nfl_fg.csv")
head(nfl_kick)
```

## Example: NFL kickers

\small

```{r, fig.width=5, fig.height=3.2}
fit_0 <- lm(Success ~ Distance, data = nfl_kick)
ggplot(data = nfl_kick, aes(Distance, Success)) + 
  geom_jitter()
```

What are the problems?



## Example: NFL kickers

\small

```{r, fig.width=6, fig.height=3.6}
fit_0 <- lm(Success ~ Distance, data = nfl_kick)
qqnorm(fit_0$resid)
```

What are the problems?



## Logistic regression model

Model:  $log(\frac{P(y = 1)}{1-P(y = 1)}) = \beta_0 + \beta_1*x_{1} + \beta_2*x_{2} + \ldots + \beta_{p-1}*x_{p-1}$

Comments: 

- Dependent variable: log-odds 
    - What are odds?
  
- Model checks more complex 
    

- Uses $z$ test statistics for parameters


## Logistic regression model

Model:  $log(\frac{P(y = 1)}{1-P(y = 1)}) = \beta_0 + \beta_1*x_{1}$

Extract probabilities:

- $P(y = 1)$:




## Estimated logistic regression model

Estimated model: $log(\frac{P(y = 1)}{1-P(y = 1)}) = \hat{\beta_0} + \hat{\beta_1}*x_{1} + \hat{\beta_2}*x_{2} + \ldots + \hat{\beta_{p-1}}*x_{p-1}$

Slope interpretation: 

- $\hat{\beta_1}$: 

- $e^{\hat{\beta_1}}$:


## Ex: Field goal kicking by distance

\footnotesize

Model: $log(\frac{P(Success = 1)}{1-P(Success = 1)}) = \beta_0 + \beta_1*Distance$


```{r}
library(broom)
fit_1 <- glm(Success ~ Distance, data = nfl_kick, family = "binomial")
tidy(fit_1)
```

Slope interpretation: $e^{\hat{\beta_1}}$

## Ex: Field goal kicking by distance

\footnotesize

```{r}
tidy(fit_1)
```

Estimate the probability of a successful 50-yard field goal:


## Ex: Field goal kicking by distance

\footnotesize

```{r}
tidy(fit_1)
```

Estimate the probability of a successful 51-yard field goal:


## Ex: Field goal kicking by distance



Use your answers on the previous slides to estimate the odds of a 51-yard field goal relative to the odds of a 50-yard field goal. Where else do you see this number?



## Model checking 

- Model checking for logistic regression relies on assessment of fit
    - Are the predicted probabilities accurate?
    - Ex: 48 to 52 yard field goals

\footnotesize

```{r}
long_FG <- filter(nfl_kick, Distance >= 48, Distance <= 52)
long_FG %>% 
  summarise(ave_success = mean(Success))
```

## Categorical predictors

\footnotesize

```{r}
fit_2 <- glm(Success ~ Distance + Grass, 
             data = nfl_kick, family = "binomial")
tidy(fit_2)
```

Estimated model


## Categorical predictors


```{r}
tidy(fit_2)
```


Slope interpretation: $e^{\hat{\beta_1}}$



## Categorical predictors


```{r}
tidy(fit_2)
```


Slope interpretation: $e^{\hat{\beta_2}}$

