---
title: 'Lecture 3: Baseball stats & Multivariate regression'
author: "Skidmore College"
output: beamer_presentation
fontsize: 10pt
---


## Multivariate regression

Model: 

$y_i = \beta_0 + \beta_1*x_{i1} + \beta_2*x_{i2} + \ldots + \beta_{p-1}*x_{i,p-1} + \epsilon_i$


Assumptions: 

- $\epsilon_i \sim N(0, \sigma^2)$
- $\epsilon_i$,$\epsilon_{i'}$ independent for all $i, i'$
- Linear relationship between $y$ and $x$

## Multivariate regression

Estimated model:  
  
$\hat{y_i} = \hat{\beta_0} + \hat{\beta_1}*x_{i1} + \hat{\beta_2}*x_{i2} + \ldots + \hat{\beta_{p-1}}*x_{i,p-1}$ 

Interpretations:

- $\hat{\beta_0}$: 
- $\hat{\beta_1}$: 


## Ex: Runs against (RA) 

```{r, message = FALSE, warning = FALSE, fig.height=2, fig.width=3}
library(tidyverse)
library(Lahman)
Teams.1 <- Teams %>% filter(yearID >= 1970)
fit.pitcher <- lm(RA ~ HRA + BBA + SOA + lgID, data = Teams.1)
```

Write the multiple regression model: 




## Ex: Runs against (RA) 

```{r}
library(broom)
tidy(fit.pitcher)  ### alternatively, use summary(fit.pitcher)
```

Write the estimated multiple regression model


## Ex: Runs against (RA) 

```{r}
tidy(fit.pitcher)
```

Interpret the slope for `SOA`. Interpret the intercept


## Ex: Runs against (RA) 

```{r}
tidy(fit.pitcher)
```

Interpret the slope for `lgID`. 


## Assumptions

Model: 

$y_i = \beta_0 + \beta_1*x_{i1} + \beta_2*x_{i2} + \ldots + \beta_{p-1}*x_{i,p-1} + \epsilon_i$


Assumptions: 

- $\epsilon_i \sim N(0, \sigma^2)$
- $\epsilon_i$,$\epsilon_{i'}$ independent for all $i, i'$
- Linear relationship between $y$ and $x$


## Ex: Runs against (RA) 

```{r, fig.height=3, fig.width=3}
qqnorm(fit.pitcher$resid)
```

## Conclusions from the model



## Open ended question 1

Write the following model, and interpret the coefficients 

```{r}
fit.offense <- lm(R ~ X2B + X3B + lgID, data = Teams.1)
tidy(fit.offense)
```


## Open ended question 2

What does the following plot say about the multiple regression model in Open Ended question 1?

```{r, fig.height=2, fig.width=3}
ggplot(Teams.1, aes(X3B, R)) + 
  geom_point() + 
  geom_smooth()
```


## Open ended question 3

What does the following plot say about the multiple regression model in Open Ended question 1?

```{r, fig.height=2, fig.width=3}
ggplot(Teams.1, aes(X2B, R)) + 
  geom_point() + 
  geom_smooth()
```


## Open ended question 4

What does the following plot say about the multiple regression model in Open Ended question 1?

```{r, fig.height=3, fig.width=3}
qqnorm(fit.offense$resid)
``` 

## Open ended question 5

Find another variable that's a significant predictor of Runs -- does it change the coefficients on the variables currently in the model?