---
title: "HW 7: NHL stats"
author: Stats and sports class
date: Fall 2020
output: 
  pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = FALSE, digits = 3, warning = FALSE)
```

# Preliminary notes for doing HW

1. All files should be knit and compiled using R Markdown. Knit early and often! I do not recommend waiting until the end of the HW to knit.

2. All questions should be answered completely, and, wherever applicable, code should be included. 

3. If you work with a partner or group, please write the names of your teammates. 

4. Copying and pasting of code is a violation of the Skidmore honor code

## Part I 

### HW Grade

Return to Homework 6 and assign yourself a grade:

- 1-3 out of 5 points: Most questions attempted, minimal effort
- 4 of 5 points: All questions attempted, complete effort, graded questions incorrect
- 4.5 of 5 points: All questions attempted, complete effort, graded questions partially correct
- 5 of 5 points: All questions attempted, graded questions perfect


## Part II: Readings

1. Read the summary model by the Evolving Wild twins: 

[https://rpubs.com/evolvingwild/395136/](https://rpubs.com/evolvingwild/395136/)

Describe five unique hockey features that were implemented in their expected goal model. That is, look through their code, and highlight various ways that hockey-specific knowledge changed how they approached the problem.

2. Compare the three variable importance plots. Which variables were more important during even-strength play? Which were more important (relatively speaking) when a team was shorthanded or at uneven strength?


## Part III: Implementation

We can access recent shot data here:

```{r}
library(tidyverse)
gitURL<- "https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/pbp_data_hockey.rds"
pbp_data <- readRDS(gzcon(url(gitURL)))
names(pbp_data)
dim(pbp_data)
```


## Question 1

Create a new variable for whether or not the shot occured during 5 v 5 play (that is, `home_skaters==5` and `away_skaters` == 5). Call this variable `is_5v5`. 

Next, identify the goal rate (e.g, how often each shot was turned into a goal) within each level of `is_5v5`. That is, were shots more or less likely to go in during 5v5 play?

## Question 2

Run the model below

```{r}
library(broom)
fit_1 <- glm(event_type == "GOAL" ~ event_distance + 
               event_angle + event_detail ,
             family = "binomial", data = pbp_data)
tidy(fit_1)
```

Interpret the coefficient on `event_detailWrist`

## Question 3

Add `is_5v5` to your model in Question 2. Using AIC criterion, identify if this creates a preferable model.

## Question 4

For `game_id == 2017020324`. For this game, use the `shot_prob` variable to estimate the total number of expected goals for each team. Next, use the `event_type` to count the number of actual goals for each team.  Did the outcome of this game match the relative shot inputs?


## Question 5

For each goal scorer (`event_player_1`), identify their total number of actual goals and their total number of expected goals. 

Look at each of the six players with the most number of expected goals. Which scored more goals than we'd expect them to? Fewer?

## Bonus

Find the one game across the two seasons of data where the difference between the observed goal differential and the actual goal differential was the biggest.
